The benefits of reporting critical effect size values

### 2024-08-20 paper/paper.Rmd
m

Despite longstanding criticisms, Null Hypothesis Significance Testing (NHST) remains a prominent approach for statistical inference in science [@cohen1994; @gigerenzer2004null; @krueger2001null]. Over the past decades, numerous proposals have emerged to enhance inference based on this approach. These include conducting power analyses to determine sample size ($n$) a priori based on theory-informed target effects, and reporting effect sizes and their confidence intervals, accompanied by pertinent related comments, rather than drawing dichotomous conclusions solely based on statistical significance [@transue2019apa]. Even for the most skeptical, there are suggestions that NHST may possess merits of its own. For instance, @wilson2020science propose that it could serve as a useful preliminary filter utilized by "original science" for screening potentially interesting effects, which should subsequently be validated by "replication science".

@@

To provide more tangible illustrations of the practical applications of critical effect size values, let us examine two real-world instances drawn from published research. The first one features a small sample, and has been targeted in the series of replications by the Open Science Collaboration [@open2015estimating], while the second one presents with a very large sample and has already attracted attention in the scientific community. The first case pertains to "Study 5" conducted by @mccrea2008self. This experiment compared two groups based on their performance percentage in a math test following a preliminary practice session with feedback of failure. One group was exposed to failure-excusing, self-handicapping thoughts, while both groups underwent the math test afterward. With a modest sample size of 28 participants split into two groups (13 and 15 participants, respectively), a directional one-tailed t-test was conducted on the percentage of correct answers to compare the groups. The resulting effect was statistically significant, $t(26) = 1.87, p (one-tailed) < .05$. Based on the reported data, Cohen's $d = 0.736$, indicating a considerable effect size. However, the critical value for Cohen's $d$ in this context was $0.645$, suggesting that, given this sample size, statistical significance is achieved only for estimated effects of substantial magnitude. Subsequent replication efforts [@open2015estimating] reaffirmed the significance of the effect, and used a somewhat larger sample ($n = 61$), $t(59)=2.325$. The estimated effect size was somehow lower than that originally reported, however, $d = 0.605$ (here, critical d = 0.428). The second scenario involves the study by @kramer2014experimental, which explored the impact of emotional content on Facebook users' experiences. With a notably large sample size of $n = 689'003$, the researchers observed a significant effect worsened emotional states when positive posts were reduced, $t(310,044) = −5.63$, $p < 0.001$ (non-directional test). Based on this and other results, the authors conclude that emotions expressed by other users on Facebook influence our own emotions. Despite the statistical significance, however, the critical values for Cohen's $d$ in this case are $-0.006$ and $+0.006$, emphasizing the importance of considering effect size beyond mere significance. This prompts reflection on whether an effect of such magnitude on such a large sample (actual $d = 0.02$) is truly meaningful in individuals' lives.

@@

With the present article, we propose that researchers compute and report the "critical effect size value(s)" in their empirical articles. This is not intended to replace other strategies aimed at enhancing the NHST approach to inference. Such strategies, such as the emphasis on estimating effect sizes with confidence intervals [@transue2019apa] or the a priori planning for statistical power are valuable in their own right. Instead, our proposal serves as a complementary tool, especially beneficial for facilitating the interpretation of results when statistical power deviates from an optimal level (typically falling below, but occasionally exceeding it). Interestingly, the critical effect size value can be retrospectively applied even to already published studies. This possible application facilitates potential reframing of the original interpretations. Serving as a tool for retrospective analysis, the critical effect size value may enable a reconsideration of the relevance of previously reported findings.

@@

Beyond enhancing study design and statistical inferences based on hypothesis tests, reporting critical effect size value(s) can also serve an educational purpose. It underscores how the distinction between a significant and non-significant result is not solely determined by the presence or absence of a true effect, but also by the sample size. By highlighting a critical effect size value, researchers can become more aware of the possibility of Type 2 errors when results are non-significant. Conversely, in studies with exceedingly large samples and in many meta-analyses, the critical effect size value(s) may serves as a reminder that any observed effect larger than a trivially small value will likely achieve significance. This emphasizes that the mere attainment of statistical significance in a test is not particularly surprising, especially in non-experimental studies.

@@

In conclusion, the reporting of the critical effect size value in empirical articles serves as a valuable addition to researchers' toolkit, aimed at augmenting transparency and facilitating the interpretability of their findings. While not designed to supplant existing practices, it provides a useful aid in interpreting newly presented and previously published results, thus advancing the understanding of research outcomes.

>>>>

Despite longstanding criticisms, Null Hypothesis Significance Testing (NHST) remains a prominent approach for statistical inference in science [@cohen1994; @gigerenzer2004null; @krueger2001null]. Over the past decades, numerous proposals have emerged to enhance inference based on this approach. These include conducting power analyses to determine sample size ($n$) \textit{a priori} based on theory-informed target effects, and reporting effect sizes and their confidence intervals, accompanied by pertinent related comments, rather than drawing dichotomous conclusions solely based on statistical significance [@transue2019apa]. Even for the most skeptical, there are suggestions that NHST may possess merits of its own. For instance, @wilson2020science propose that it could serve as a useful preliminary filter utilized by "original science" for screening potentially interesting effects, which should subsequently be validated by "replication science".

@@

To provide more tangible illustrations of the practical applications of critical effect size values, let us examine two real-world instances drawn from published research. The first one features a small sample and has been targeted in the series of replications by the Open Science Collaboration [@open2015estimating], while the second one presents with a very large sample and has already attracted attention in the scientific community. The first case pertains to "Study 5" conducted by @mccrea2008self. This experiment compared two groups based on their performance percentage in a math test following a preliminary practice session with feedback of failure. One group was exposed to failure-excusing, self-handicapping thoughts, while both groups underwent the math test afterward. With a modest sample size of 28 participants split into two groups (13 and 15 participants, respectively), a directional one-tailed t-test was conducted on the percentage of correct answers to compare the groups. The resulting effect was statistically significant, $t(26) = 1.87$, $p$ (one-tailed) $< .05$. Based on the reported data, Cohen's $d = 0.736$, indicating a considerable effect size. However, the critical value for Cohen's $d$ in this context was $0.645$, suggesting that, given this sample size, statistical significance is achieved only for estimated effects of substantial magnitude. Subsequent replication efforts [@open2015estimating] reaffirmed the significance of the effect, and used a somewhat larger sample ($n = 61$), $t(59)=2.325$. The estimated effect size was somehow lower than that originally reported, however, $d = 0.605$ (here, critical d = 0.428). The second scenario involves the study by @kramer2014experimental, which explored the impact of emotional content on Facebook users' experiences. With a notably large sample size of $n = 689'003$, the researchers observed a significant effect worsened emotional states when positive posts were reduced, $t(310,044) = −5.63$, $p < 0.001$ (non-directional test). Based on this and other results, the authors conclude that emotions expressed by other users on Facebook influence our own emotions. Despite the statistical significance, however, the critical values for Cohen's $d$ in this case are $-0.006$ and $+0.006$, emphasizing the importance of considering effect size beyond mere significance. This prompts reflection on whether an effect of such magnitude on such a large sample (actual $d = 0.02$) is truly meaningful in individuals' lives.

@@

With the present article, we propose that researchers compute and report the "critical effect size value(s)" in their empirical articles. This is not intended to replace other strategies aimed at enhancing the NHST approach to inference. Such strategies, such as the emphasis on estimating effect sizes with confidence intervals [@transue2019apa] or the \textit{a priori} planning for statistical power are valuable in their own right. Instead, our proposal serves as a complementary tool, especially beneficial for facilitating the interpretation of results when statistical power deviates from an optimal level (typically falling below, but occasionally exceeding it). Interestingly, the critical effect size value can be retrospectively applied even to already published studies. This possible application facilitates potential reframing of the original interpretations. Serving as a tool for retrospective analysis, critical effect size values may enable a reconsideration of the relevance of previously reported findings.

@@

Beyond enhancing study design and statistical inferences based on hypothesis tests, reporting critical effect size value(s) can also serve an educational purpose. It underscores how the distinction between a significant and non-significant result is not solely determined by the presence or absence of a true effect, but also by the sample size. By highlighting a critical effect size value, researchers can become more aware of the possibility of Type 2 errors when results are non-significant. Conversely, in studies with exceedingly large samples and in many meta-analyses, the critical effect size value(s) may serve as a reminder that any observed effect larger than a trivially small value will likely achieve significance. This emphasizes that the mere attainment of statistical significance in a test is not particularly surprising, especially in non-experimental studies.

@@

In conclusion, reporting critical effect size values in empirical articles serves as a valuable addition to researchers' toolkit, aimed at augmenting transparency and facilitating the interpretability of their findings. While not designed to supplant existing practices, it provides a useful aid in interpreting newly presented and previously published results, thus advancing the understanding of research outcomes.

### 2024-08-20 paper/paper.tex
m

Despite longstanding criticisms, Null Hypothesis Significance Testing (NHST) remains a prominent approach for statistical inference in science (Cohen, 1994; Gigerenzer et al., 2004; Krueger, 2001). Over the past decades, numerous proposals have emerged to enhance inference based on this approach. These include conducting power analyses to determine sample size (\(n\)) a priori based on theory-informed target effects, and reporting effect sizes and their confidence intervals, accompanied by pertinent related comments, rather than drawing dichotomous conclusions solely based on statistical significance (Transue, 2019). Even for the most skeptical, there are suggestions that NHST may possess merits of its own. For instance, Wilson et al. (2020) propose that it could serve as a useful preliminary filter utilized by ``original science'' for screening potentially interesting effects, which should subsequently be validated by ``replication science''.

@@

To provide more tangible illustrations of the practical applications of critical effect size values, let us examine two real-world instances drawn from published research. The first one features a small sample, and has been targeted in the series of replications by the Open Science Collaboration (Collaboration, 2015), while the second one presents with a very large sample and has already attracted attention in the scientific community. The first case pertains to ``Study 5'' conducted by McCrea (2008). This experiment compared two groups based on their performance percentage in a math test following a preliminary practice session with feedback of failure. One group was exposed to failure-excusing, self-handicapping thoughts, while both groups underwent the math test afterward. With a modest sample size of 28 participants split into two groups (13 and 15 participants, respectively), a directional one-tailed t-test was conducted on the percentage of correct answers to compare the groups. The resulting effect was statistically significant, \(t(26) = 1.87, p (one-tailed) < .05\). Based on the reported data, Cohen's \(d = 0.736\), indicating a considerable effect size. However, the critical value for Cohen's \(d\) in this context was \(0.645\), suggesting that, given this sample size, statistical significance is achieved only for estimated effects of substantial magnitude. Subsequent replication efforts (Collaboration, 2015) reaffirmed the significance of the effect, and used a somewhat larger sample (\(n = 61\)), \(t(59)=2.325\). The estimated effect size was somehow lower than that originally reported, however, \(d = 0.605\) (here, critical d = 0.428). The second scenario involves the study by Kramer et al. (2014), which explored the impact of emotional content on Facebook users' experiences. With a notably large sample size of \(n = 689'003\), the researchers observed a significant effect worsened emotional states when positive posts were reduced, \(t(310,044) = −5.63\), \(p < 0.001\) (non-directional test). Based on this and other results, the authors conclude that emotions expressed by other users on Facebook influence our own emotions. Despite the statistical significance, however, the critical values for Cohen's \(d\) in this case are \(-0.006\) and \(+0.006\), emphasizing the importance of considering effect size beyond mere significance. This prompts reflection on whether an effect of such magnitude on such a large sample (actual \(d = 0.02\)) is truly meaningful in individuals' lives.

@@

With the present article, we propose that researchers compute and report the ``critical effect size value(s)'' in their empirical articles. This is not intended to replace other strategies aimed at enhancing the NHST approach to inference. Such strategies, such as the emphasis on estimating effect sizes with confidence intervals (Transue, 2019) or the a priori planning for statistical power are valuable in their own right. Instead, our proposal serves as a complementary tool, especially beneficial for facilitating the interpretation of results when statistical power deviates from an optimal level (typically falling below, but occasionally exceeding it). Interestingly, the critical effect size value can be retrospectively applied even to already published studies. This possible application facilitates potential reframing of the original interpretations. Serving as a tool for retrospective analysis, the critical effect size value may enable a reconsideration of the relevance of previously reported findings.

>>>>

Despite longstanding criticisms, Null Hypothesis Significance Testing (NHST) remains a prominent approach for statistical inference in science (Cohen, 1994; Gigerenzer et al., 2004; Krueger, 2001). Over the past decades, numerous proposals have emerged to enhance inference based on this approach. These include conducting power analyses to determine sample size (\(n\)) \textit{a priori} based on theory-informed target effects, and reporting effect sizes and their confidence intervals, accompanied by pertinent related comments, rather than drawing dichotomous conclusions solely based on statistical significance (Transue, 2019). Even for the most skeptical, there are suggestions that NHST may possess merits of its own. For instance, Wilson et al. (2020) propose that it could serve as a useful preliminary filter utilized by ``original science'' for screening potentially interesting effects, which should subsequently be validated by ``replication science''.

@@

To provide more tangible illustrations of the practical applications of critical effect size values, let us examine two real-world instances drawn from published research. The first one features a small sample and has been targeted in the series of replications by the Open Science Collaboration (Collaboration, 2015), while the second one presents with a very large sample and has already attracted attention in the scientific community. The first case pertains to ``Study 5'' conducted by McCrea (2008). This experiment compared two groups based on their performance percentage in a math test following a preliminary practice session with feedback of failure. One group was exposed to failure-excusing, self-handicapping thoughts, while both groups underwent the math test afterward. With a modest sample size of 28 participants split into two groups (13 and 15 participants, respectively), a directional one-tailed t-test was conducted on the percentage of correct answers to compare the groups. The resulting effect was statistically significant, \(t(26) = 1.87\), \(p\) (one-tailed) \(< .05\). Based on the reported data, Cohen's \(d = 0.736\), indicating a considerable effect size. However, the critical value for Cohen's \(d\) in this context was \(0.645\), suggesting that, given this sample size, statistical significance is achieved only for estimated effects of substantial magnitude. Subsequent replication efforts (Collaboration, 2015) reaffirmed the significance of the effect, and used a somewhat larger sample (\(n = 61\)), \(t(59)=2.325\). The estimated effect size was somehow lower than that originally reported, however, \(d = 0.605\) (here, critical d = 0.428). The second scenario involves the study by Kramer et al. (2014), which explored the impact of emotional content on Facebook users' experiences. With a notably large sample size of \(n = 689'003\), the researchers observed a significant effect worsened emotional states when positive posts were reduced, \(t(310,044) = −5.63\), \(p < 0.001\) (non-directional test). Based on this and other results, the authors conclude that emotions expressed by other users on Facebook influence our own emotions. Despite the statistical significance, however, the critical values for Cohen's \(d\) in this case are \(-0.006\) and \(+0.006\), emphasizing the importance of considering effect size beyond mere significance. This prompts reflection on whether an effect of such magnitude on such a large sample (actual \(d = 0.02\)) is truly meaningful in individuals' lives.

@@

With the present article, we propose that researchers compute and report the ``critical effect size value(s)'' in their empirical articles. This is not intended to replace other strategies aimed at enhancing the NHST approach to inference. Such strategies, such as the emphasis on estimating effect sizes with confidence intervals (Transue, 2019) or the \textit{a priori} planning for statistical power are valuable in their own right. Instead, our proposal serves as a complementary tool, especially beneficial for facilitating the interpretation of results when statistical power deviates from an optimal level (typically falling below, but occasionally exceeding it). Interestingly, the critical effect size value can be retrospectively applied even to already published studies. This possible application facilitates potential reframing of the original interpretations. Serving as a tool for retrospective analysis, critical effect size values may enable a reconsideration of the relevance of previously reported findings.
